Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2021-12-22 manuscripts/owp_fim4_2021/sections/abstract_plainTextSummary.tex
Manuscript edits made after feedback from co-authors

\begin{abstract}
The National Water Model Version 2.1 (NWM V2.1) seeks to expand upon existing systems to provide hourly forecast discharges at four time horizons (1, 3, 10, and 30 days) at over 2.7 million locations throughout the continental United States, Hawaii and Puerto Rico.
Additional modeling capabilities are required to convert these 1-D discharge forecasts into 2-D fluvial inundation extents used for forecasting purposes to protect life and property.
Height Above Nearest Drainage (HAND) coupled with the use of synthetic rating curves can convert discharges into stages and stages to inundation extents in a rapid manner across large spatial domains and at high resolutions.
The Office of Water Prediction Flood Inundation Mapping (OWP FIM) model `Cahaba' is an enhancement of the HAND method developed to improve upon some of HAND's limitations.
Among the implemented enhancements are efficient computational performance, support for Hawaii and Puerto Rico NWM domains, enforcement of levee data, support for the latest digital elevation model and stream network data sources, advanced stream burning techniques, thalweg conditioning, and simplified Manning's N variation.
Most importantly, we detail an issue previously mentioned with HAND as its FIM skill is dependent on a drainage threshold parameter used to delineate stream lines.
Since our modeling assumptions are different, we illustrate how reducing Horton-Strahler stream order of the stream networks used within each HAND processing unit down to unary can improve FIM skill.
This is accomplished by deriving level paths for the National Water Model stream network and utilizing those definitions for deriving HAND datasets independently from other level paths.
`Cahaba' was evaluated at 49 Base Level Engineering (BLE) sites furnished by the Federal Emergency Management Agency (FEMA).
The modeled discharges from the BLE sections were used for the 100 and 500 year flood events as input to OWP FIM instead of NWM discharges to isolate out factors related to hydro-meteorology.
The resulting maps are compared to the modeled extents from the BLE and binary contingency statistics are computed.
Across the 49 sites, Critical Success Index (CSI) was found to increase across FIM versions driven by improvements to the probability of detection (POD) and slight reduction in False Alarm Ratio (FAR).
Each successive version of HAND here utilizes greater stream order reduction illustrating a key limitation of HAND and enhanced agreement when utilizing stream networks of lower stream orders as a datum for HAND.
Additionally we detail the effects of two different Manning's N values and how order reduction drives rating curve bias downward due to the nature of reach averaged synthetic rating curves.
These capabilities were implemented in an open-source package available on GitHub (\url{https://github.com/NOAA-OWP/cahaba}) for continued use and collaboration with the research community.
%
\end{abstract}
%
\section*{Plain Language Summary}
Flooding is one of the most impactful natural disasters on life and property.
The United States National Water Model (NWM) seeks to provide flood forecasts for the entire country so that adequate warnings can be raised to the public to enable safe evacuations and protective measures.
In order to convert forecasts flow rates from the NWM to inundation extents, a model is used that converts elevations relative to sea-level to elevations relative to the nearest river bottom elevation.
This model known as Height Above Nearest Drainage (HAND) suffers from issues in mapping performance where rivers converge.
The maps were compared to other, more rigorous models and shown to perform well when compared to previous versions given certain enhancements to the way HAND to computed.
%

>>>>

\begin{abstract}
The National Water Model Version 2.1 (NWM V2.1) seeks to expand upon existing systems to provide hourly forecast discharges at four time horizons (1, 3, 10, and 30 days) at over 2.7 million locations throughout the continental United States, Hawaii and Puerto Rico.
Additional modeling capabilities are required to convert these 1-D discharge forecasts into 2-D fluvial inundation extents used for forecasting purposes to protect life and property.
Height Above Nearest Drainage (HAND) coupled with the use of synthetic rating curves can convert discharges into stages and stages to inundation extents in a rapid manner across large spatial domains and at high resolutions.
The Office of Water Prediction Flood Inundation Mapping (OWP FIM) model `Cahaba' is an enhancement of the HAND method developed to improve upon some of HAND's limitations.
Among the implemented enhancements are efficient computational performance, support for Hawaii and Puerto Rico NWM domains, enforcement of levee elevation data, support for the latest digital elevation model and stream network data sources, advanced stream burning techniques, thalweg conditioning, and simplified Manning's N variation.
Most importantly, we detail an issue previously mentioned with HAND as its FIM skill is dependent on a drainage threshold parameter used to delineate stream lines.
Since our modeling assumptions are different, we illustrate how reducing Horton-Strahler stream order of the stream networks used within each HAND processing unit down to unary can improve FIM skill.
This is accomplished by deriving level paths for the National Water Model stream network and utilizing those definitions for deriving HAND datasets independently from other level paths.
`Cahaba' was evaluated at 49 Base Level Engineering (BLE) sites furnished by the Federal Emergency Management Agency (FEMA).
The modeled discharges from the BLE sections were used for the 100 and 500 year flood events as input to OWP FIM instead of NWM discharges to isolate out factors related to hydro-meteorology.
The resulting maps are compared to the modeled extents from the BLE and binary contingency statistics are computed.
Across the 49 sites, Critical Success Index (CSI) was found to increase across FIM versions driven by improvements to the Probability of Detection (POD) and slight reduction in False Alarm Ratio (FAR).
Each successive version of HAND here utilizes greater stream order reduction illustrating a key limitation of HAND and enhanced agreement when utilizing stream networks of lower stream orders as a datum for HAND.
Additionally we detail the effects of two different Manning's N values and how order reduction drives rating curve bias downward due to the nature of reach averaged synthetic rating curves.
These capabilities were implemented in an open-source package available on GitHub (\url{https://github.com/NOAA-OWP/cahaba}) for continued use and collaboration with the research community.
%
\end{abstract}
%
\section*{Plain Language Summary}
Flooding is one of the most impactful natural disasters on life and property.
The United States National Water Model (NWM) seeks to provide flood forecasts for the entire country so that adequate warnings can be raised to the public to enable safe evacuations and protective measures.
In order to convert forecasted flow rates from the NWM to flood inundation maps (FIM), a model is used that converts elevation data from height above mean sea-level to height above the nearest river bottom.
This model known as Height Above Nearest Drainage (HAND) suffers from issues in mapping performance where rivers converge.
We developed a technique that mitigates these errors by removing consideration for neighboring tributaries in the relative elevation computation process.
We compared these HAND derived FIMs to maps from more realistic models and found and improvement in mapping performance.
%

### 2021-12-22 manuscripts/owp_fim4_2021/sections/discussion.tex
Manuscript edits made after feedback from co-authors

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's N values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V[y] or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B[y] or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively. 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significant higher values than MS which has higher values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but since both are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ is plotted against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments than encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

For future development, we assert using experience from qualitative analysis that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM.
The bathymetry of the NED 10m DEM is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's N which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3m and 1m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

Lastly, after errors introduced by conflation, poor roughness estimation, bathymetric/elevation adjustment are accounted for, HAND still has another fundamental limitation that is inherently baked into how it works.
For HAND to be derived and thus create a FIM for a given area, that area must entirely drain to the stream network and the stream network must also drain itself.
In other words, an entire area eligible for flooding must monotonically decrease in elevation. 
DEM's naturally don't do this and the dynamics of true flood events don't follow drainage patterns.
Enforcing this assumption for HAND leads to significant amount of DEM manipulations that introduce basic errors.
These errors are deep into the assumptions of HAND and thus more difficult to disentangle.
Ultimately, the use of more advanced 2-D hydrodynamic models should be considered for dealing with this limitation of HAND but would come at significant expense at the given high resolution across very large scales.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/rating_curve_comparison.jpg}
\caption{Illustrates average quantities for the three methods, FR, MS, and GMS, for each stage value (m). 
The values are (a) Discharge $m^3s^{-1}$, (b) Volume $m^3$, (c) Bed Area $m^2$, (d) a function of Volume and Bed Area, and (e) number of pixels.
}
\label{fig:rating_curve_comparison}
\end{figure}
%
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/gms_enhancement.jpg}
\caption{OWP FIM `Cahaba' inundation agreement, TP, FP, FN, and TN, with BLE HEC-RAS maps in HUC 11140105.
Catchment boundaries and stream lines are shown in white and dotted green, respectively.
Sub-figure (a) shows agreement of FR HAND denoting significant areas of under-prediction due to junctions and catchment boundaries.
Meanwhile, (b) shows the agreement for GMS and much larger catchments leading to much better inundation agreement for this given reach. 
Overall, this illustrates the benefits of stream order reduction for deriving HAND datasets.
}
\label{fig:gms_enhancement}
\end{figure}
%

>>>>

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's N values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V[y] or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B[y] or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively. 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but since both are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ is plotted against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

Our qualitative analysis suggests that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the NED 10m DEM is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's N which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3m and 1m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

Lastly, after errors introduced by conflation, poor roughness estimation, bathymetric/elevation adjustment are accounted for, HAND still has another fundamental limitation that is inherently baked into how it works.
For HAND to be derived and thus create a FIM for a given area, that area must entirely drain to the stream network and the stream network must also drain itself.
In other words, an entire area eligible for flooding must monotonically decrease in elevation. 
DEM's naturally don't do this and the dynamics of true flood events don't follow drainage patterns.
Enforcing this assumption for HAND leads to significant amount of DEM manipulations that introduce basic errors.
These errors are deep into the assumptions of HAND and thus more difficult to disentangle.
Ultimately, the use of more advanced 2-D hydrodynamic models should be considered for dealing with this limitation of HAND but would come at significant expense at the given high resolution across very large spatial scales and frequent forecast resolutions.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/rating_curve_comparison.jpg}
\caption{Illustrates average quantities for the three methods, FR, MS, and GMS, for each stage value (m). 
The values are (a) Discharge $m^3s^{-1}$, (b) Volume $m^3$, (c) Bed Area $m^2$, (d) a function of Volume and Bed Area, and (e) number of pixels.
}
\label{fig:rating_curve_comparison}
\end{figure}
%
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/gms_enhancement.jpg}
\caption{OWP FIM `Cahaba' inundation agreement, TP, FP, FN, and TN, with BLE HEC-RAS maps in HUC 11140105.
Catchment boundaries and stream lines are shown in white and dotted green, respectively.
Sub-figure (a) shows agreement of FR HAND denoting significant areas of under-prediction due to junctions and catchment boundaries.
Meanwhile, (b) shows the agreement for GMS and much larger catchments leading to much better inundation agreement for this given reach. 
Overall, this illustrates the benefits of stream order reduction for deriving HAND datasets.
}
\label{fig:gms_enhancement}
\end{figure}
%

### 2021-12-22 manuscripts/owp_fim4_2021/sections/introduction.tex
Manuscript edits made after feedback from co-authors

\caption{Forecast points with and without FIM in United States' Advanced Hydrologic Prediction System. Also show are the National Water Model stream networks at the full resolution (FR) and Replace and Reroute (RnR) Mainstems (MS) resolution.}

@@

The term hydrofabric is used within the NWM jargon to describe the subset of hydrography comprised of geospatial datasets required for hydrologic modeling including but not limited to stream networks, catchments, channel properties, and elevation data. 

@@

HAND normalizes topography along the nearest drainage path and its been demonstrated to be a good proxy and indicator of a series of important environmental conditions including soil environments, landscape classes, soil gravitational potentials, geomorphologies, soil moisture, and ground water dynamics \cite{renno2008hand,nobre2011height}. 

@@

The terrain index as even gone on to provide additional utility in the observation of riverine flood inundation mapping from remote sensing especially in areas of high electromagnetic interference such as vegetated and anthropogenic areas \cite{aristizabal2020high,shastry2019using,huang2017comparison,twele2016sentinel}.
\citeA{zheng2018river} developed methodology for determining stage-discharge relationships known as synthetic rating curves (SRC) by sampling reach-averaged parameters from HAND datasets and inputting into the Manning's equation \cite{gauckler1867etudes,manning1890flow}.
This collection of methods, coupling HAND with SRCs, has been experimented and compared to other sources of FIM including engineering scale models, in-situ observation, and remote sensing based observation with solid results in large spatial scale applications \cite{godbout2019error,johnson2019integrated,garousi2019terrain,nobre2016hand,afshari2018comparison,zheng2018geoflood,teng2015rapid,teng2017flood,zhang2018comparative}.

@@

Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEM's such as the National Elevation Dataset (NED) at the 10m scale, HAND has been implemented into software for large-scale, continental computation. 

@@

The original NFIE dataset was employed by the NWC to produce forecast FIM from the NWM for use within its network of RFC's for additional guidance in hydro-blind regions.

@@

Related to these, the USGS has invested in relative elevation HAND like methods via work in the GIS Flood Tool (GFT) that also uses synthetic rating curves with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OWP FIM `Cahaba'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Many of those in the academic community assessing HAND's efficacy for producing FIM have noted opportunities for improvement. 
\citeA{godbout2019error} found how reach length and slope are important parameters for maximizing mapping skill with the moderate values performing best. 
The co-linearity of reach length and slope led \citeA{godbout2019error} to propose that reaches of extreme lengths performed worse because of the extreme slope values, a parameter directly represented in Manning's equation. 
Issues with the reach-average approaches have been documented in \citeA{tuozzolo2019impact} where large within reach variance of the roughness Manning's n coefficient have been observed.
Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).
Use of higher resolution DEMs in that study was motivated by previous work with Lidar DEMs and least-cost thalweg derivations \cite{zheng2018geoflood}.
Further work by \citeA{johnson2019integrated} noted the general under-prediction of HAND and suggested tuning the Manning's n parameter to better align SRC's with observations. 
Additionally, the sensitivity to low topographic relief and channel slope have been observed \cite{johnson2019integrated,godbout2019error}. 
Most notably, HAND has been shown to demonstrate sensitivity to drainage network density known colloquially as the catchment boundary problem \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is noted at higher order streams with large flows and low Froude numbers and is manifested by a constriction in the inundation extents at the junction location.
These findings suggest improvements to HAND are required that utilize advanced computational algorithms and software to compute a FIM hydrofabric required for producing continental-scale FIM.

With all of the latest developments in the realm of continental FIM (CFIM) noted in the previous paragraph, a fast, open-source framework for accelerating the research to operational pipeline is required.
Here we introduce OWP FIM `Cahaba' that utilizes a few of the latest techniques in HAND based FIM oriented for use with NWM in continental scale operational forecasting settings. 
This framework is open-source and modular enabling and accelerating the research to operational development pipeline.
Automated evaluation tooling and processed test case data enable for the rapid testing and evaluation of new techniques in consistent contexts.
In addition to developed tooling, we introduce research demonstrating how FIM performance skill with HAND can be improved by reducing Horton-Strahler stream orders \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} of the stream networks.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to stream threshold which serves as a proxy for stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths, can enhance FIM skill.
The following methods and results describe the work in more detail and demonstrates its efficacy in producing enhanced FIM for the NWM with applications.
%

>>>>

\caption{Forecast points with and without FIM in United States' Advanced Hydrologic Prediction System. Also show are the National Water Model stream networks at the full resolution (FR) and Replace and Route (RnR) Mainstems (MS) resolution.}

@@

The term hydrofabric is used within the NWM jargon to describe the subset of hydrography composed of the geospatial datasets required for hydrologic modeling including but not limited to stream networks, catchments, channel properties, and elevation data. 

@@

HAND normalizes topography along the nearest drainage path and its been demonstrated to be a good proxy and indicator of a series of important environmental conditions including soil environments, landscape classes, soil gravitational potentials, geomorphologies, soil moisture, and groundwater dynamics \cite{renno2008hand,nobre2011height}. 

@@

The terrain index has even gone on to provide additional utility in the observation of riverine flood inundation mapping from remote sensing especially in areas of high electromagnetic interference such as vegetated and anthropogenic areas \cite{aristizabal2020high,shastry2019using,huang2017comparison,twele2016sentinel}.
\citeA{zheng2018river} developed a methodology for determining stage-discharge relationships known as synthetic rating curves (SRC) by sampling reach-averaged parameters from HAND datasets and inputting into the Manning's equation \cite{gauckler1867etudes,manning1890flow}.
This collection of methods, coupling HAND with SRCs, have been experimented with and compared to other sources of FIM including engineering scale models, in-situ observation, and remote sensing based observation with solid results in large spatial scale applications \cite{godbout2019error,johnson2019integrated,garousi2019terrain,nobre2016hand,afshari2018comparison,zheng2018geoflood,teng2015rapid,teng2017flood,zhang2018comparative}.

@@

Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the National Elevation Dataset (NED) at the 10m scale, HAND has been implemented into software for large-scale, continental computation. 

@@

The original NFIE dataset was employed by the NWC to produce forecast FIM from the NWM for use within its network of RFCs for additional guidance in hydro-blind regions.

@@

Related to these, the USGS has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses synthetic rating curves with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{OWP FIM `Cahaba'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Many of those in the academic community assessing HAND's efficacy for producing FIM have noted opportunities for improvement. 
\citeA{godbout2019error} found how reach length and slope are important parameters for maximizing mapping skill with the moderate values performing best. 
The co-linearity of reach length and slope led \citeA{godbout2019error} to propose that reaches of extreme lengths performed worse because of the extreme slope values, a parameter directly represented in Manning's equation. 
Issues with the reach-average approaches have been documented in \citeA{tuozzolo2019impact} where large within reach variance of the roughness Manning's n coefficient have been observed.
Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).
The use of higher resolution DEMs in that study was motivated by previous work with Lidar DEMs and least-cost thalweg derivations \cite{zheng2018geoflood}.
Further work by \citeA{johnson2019integrated} noted the general under-prediction of HAND and suggested tuning the Manning's n parameter to better align SRC's with observations. 
Additionally, the sensitivity to low topographic relief and channel slope have been observed \cite{johnson2019integrated,godbout2019error}. 
Most notably, HAND has been shown to demonstrate sensitivity to drainage network density known colloquially as the catchment boundary problem \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is noted at higher order streams with large flows and low Froude numbers and is manifested by a constriction in the inundation extents at the junction location.
These findings suggest improvements to HAND are required that utilize advanced computational algorithms and software to compute a FIM hydrofabric required for producing continental-scale FIM.

With all of the latest developments in the realm of continental FIM (CFIM) noted in the previous paragraph, a fast, open-source framework for accelerating the research to operations pipeline is required.
Here we introduce OWP FIM `Cahaba' that utilizes a few of the latest techniques in HAND based FIM oriented for use with NWM in continental scale operational forecasting settings. 
This framework is open-source and modular enabling and accelerating the research to operational development pipeline.
Automated evaluation tools and processed test case data enable the rapid testing and evaluation of new techniques in consistent contexts.
In addition to developed tooling, we introduce research demonstrating how FIM performance skill with HAND can be improved by reducing Horton-Strahler stream orders \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} of the stream networks.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to stream threshold which serves as a proxy for stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths, can enhance FIM skill.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%
